---
title: "Characterization of African *P. malariae* clusters"
author: "Margaux Lefebvre"
date: "`r Sys.Date()`"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE) #prevent all chunks from running
library(tidyverse)
```

# PCA and ancestry plots

## ANGSD input and LD-pruning

Version: angsd v0.940, python v3.8.12, R v4.5.2, ngsLD v1.2.0.
```{bash}
listbam=list_bams_Africa.txt #all samples without outgroup
REF=GCF_900090045.1_PmUG01_genomic.fna
CORE=core_malariae.txt

# Create Beagle input for the pruning
angsd -b $listbam -ref $REF -out PCA_coregenome.Africa.Africa \
        -rf $CORE -uniqueOnly 1 \
        -minMapQ 20 -minQ 20 -doCounts 1 \
        -GL 2 -doGlf 2 -nThreads 8 -doMajorMinor 1 -SNP_pval 1e-6 -doMaf 1 -minMaf 0.01

# Prepare a pos file with the mafs filre
zcat PCA_coregenome.Africa.Africa.mafs.gz | cut -f 1,2 |  sed 's/:/_/g'| gzip > PCA_coregenome.Africa.pos.gz

# Run ngsLD
ngsLD \
--geno PCA_coregenome.Africa.beagle.gz \
--posH PCA_coregenome.Africa.pos.gz \
--probs \
--n_ind 160 \
--n_sites 260409 \
--max_kb_dist 1 \
--n_threads 8 \
--out PCA_coregenome.Africa.ld

# Run prune_graph
prune_graph --header \
--in PCA_coregenome.Africa.ld --weight-field "r2" \
--weight-filter "dist <= 5000 && r2 >= 0.5" \
--out PCA_coregenome.Africa.snp.unlinked.id
```

Generate an LD-pruned SNP list:

```{r}
pruned_position <- as.integer(gsub(paste0("NC_0417[0-9][0-9].1:"), "", readLines(paste0("./Data/PCA_coregenome.Africa.snp.unlinked.id"))))
snp_list <- read.table(paste0("./Data/PCA_coregenome.Africa.mafs.gz"), stringsAsFactors = F, header = T)[,1:4]
pruned_snp_list <- snp_list[snp_list$position %in% pruned_position, ]
write.table(pruned_snp_list, paste0("./Data/PCA_coregenome.Africa.snp.LDpruned.list"), col.names = F, row.names = F, quote = F, sep = "\t")
```

Create the input for PCAngsd
```{bash}
listbam=list_bams_Africa.txt #all samples without outgroup
REF=GCF_900090045.1_PmUG01_genomic.fna
CORE=core_malariae.txt

# Create input
angsd sites index PCA_coregenome.Africa.snp.LDpruned.list # mandatory

# Create Beagle input for the pruning
angsd -b $listbam -ref $REF -out PCA_coregenome.Africa_final \
        -rf $CORE -uniqueOnly 1 \
        -minMapQ 20 -minQ 20 -doCounts 1 \
        -GL 2 -doGlf 2 -nThreads 8 -doMajorMinor 1 -doMAF 1 -SNP_pval 1e-6 \
        -doPost 1 -doIBS 1 -doCov 1 -makeMatrix 1 -sites PCA_coregenome.Africa.snp.LDpruned.list
```

## PCA

Version: PCAngsd v1.36.1.
```{bash}
pcangsd -b PCA_coregenome.Africa_final.beagle.gz -o PCA.Africa_final.mat --iter 500 -t 8 --maf 0.01
```

## Ancestry plots

Version: PCAngsd v1.36.1.
```{bash}
# Run PCAngsd for ancestry plots
pcangsd -b PCA_coregenome.Africa_final.beagle.gz --admix --admix-K 2 -o Africa.ancestry --iter 500 -t 8 --maf 0.01
```

Plot on the map
```{r}
library(tidyverse)
Samples_data<-read_delim("metadata_final.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)

indiv_order<-read_delim("./Data/list_samples_Africa.txt", 
    delim = "\t", escape_double = FALSE, 
    col_names = "sample_ID", trim_ws = TRUE)

info<-inner_join(indiv_order,Samples_data, join_by(sample_ID == Name))

Qmat <- read_table("./Data/Africa.ancestry.admix.2.Q", 
    col_names = c("Q1","Q2"))
total_map<-cbind(info, Qmat)
total_map[,c(26:27)]<-total_map[,c(26:27)]*100
total_map$Longitude<-as.numeric(total_map$Longitude)
mean_ancestry<-total_map %>% 
  group_by(Country) %>%
  summarise(Q1 = mean(Q1),
            Q2 = mean(Q2), Lattitude=country_Lat, Longitude=country_Long)
mean_ancestry<-unique(mean_ancestry)

# Libraries
library(ggplot2)
library(dplyr)
library(scatterpie)
library(ggnewscale)
library(ggrepel)
 
# Get the world polygon and extract UK
library(maps)
world <- map_data("world")

mapplot1<-ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group), fill="white",col="grey") +
  theme_void() + coord_map() +coord_cartesian(xlim = c(-20,50), ylim = c(-35,35))

mapplot2<-mapplot1  +
geom_scatterpie(data=mean_ancestry, aes(x=Longitude, y=Lattitude, r= 1.2),cols = colnames(mean_ancestry[,c(2:3)]), alpha=01, color=NA) +scale_fill_manual(values = c('#e6194b', '#ffe119'))

mapplot2+ theme_bw()+ theme(legend.position = "none")+labs(title = paste0("Ancestry K=2 in Africa"))+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

# Genome-scans

For the genome scans, I kept the African samples with more than 70% ancestry in one African cluster and Thai samples as outgroup population.

## Create the VCF file

Version: bcftools v1.16.
```{bash}
# First create a VCF with all sites
bcftools mpileup -f GCF_900090045.1_PmUG01_genomic.fna -b list_bams_clusters.txt -R core_malariae.bcf.txt -q 20 -Q 20 | bcftools call -m -Oz -f GQ -o clusters_samples.allsites.vcf.gz
tabix clusters_samples.allsites.vcf.gz

bcftools view -M2 -i 'DP>10 & F_MISSING<1' clusters_samples.allsites.vcf.gz -Oz -o clusters_samples.filtered.vcf.gz
tabix clusters_samples.filtered.vcf.gz
```

## Nucleotide diversity and Tajima's D

Version: pixy v2.0.0.beta14, R v4.5.2.

Calculate pi and Tajima's D with pixy with non-overlapping 500pb windows.
```{bash}
pixy --stats pi tajima_d \
--vcf clusters_samples.filtered.vcf.gz \
--populations clusters_popfile.txt \
--window_size 500 \
--n_cores 16
```

Check if pi or Tajima's D are significantly different (the script is only for pi but it's the same for Tajima's D).
```{r}
library(readr)
library(qqplotr)
library(MASS)
library(rstatix)

pixy_pi <- read_delim("./Data/pixy_pi.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)

pixy_pi <- na.omit(pixy_pi)
pixy_pi<-subset(pixy_pi, no_sites>=100) # Remove windows with less than 100 pb
pixy_pi<-subset(pixy_pi, pop!="Thai") # Remove the Thai population because not of interest for this analysis

# Check if distribution are normal with QQ plot
pixy_pi %>%
  ggplot(aes(sample = avg_pi)) +
     stat_qq_band() +
 stat_qq_line() +
 stat_qq_point()+ labs(x = "Theoretical Quantiles", y = "Sample Quantiles")+ ggtitle("QQ plot for \u03c0 values")+
  facet_wrap(~pop, scales = "free_y")+theme_bw()

# Test if significantly different with Wilcox test
test<-pixy_pi %>% 
  wilcox_test(avg_pi ~ pop, p.adjust.method = "none") # no adjustment because we have only one comparison
# Check the mean of each cluster
pixy_pi %>%
  group_by(pop)%>%
  summarise(mean = mean(avg_pi))
```

## *F<sub>ST</sub>* and *D<sub>XY</sub>*

Version: pixy v2.0.0.beta14, R v4.5.2, bedtools v2.31.1.

Calculate *F<sub>ST</sub>* and *D<sub>XY</sub>* with pixy with sliding windows of 5kb and 500pb steps.
```{bash}
pixy --stats fst dxy \
--vcf clusters_samples.filtered.vcf.gz \
--populations clusters_popfile.txt \
--bed_file slid.500stp.5kbwin.bed \
--n_cores 16
# the bed file corresponds to coordinates for sliding windows of 5kb and 500pb steps
```

The genome-scans for seeking differentiation islands.
```{r}
library(tidyverse)
library(readr)
library(dplyr)

# Load FST data
pixy_fst <- read_delim("./Data/win_slid/pixy_fst.txt", delim="\t", trim_ws=TRUE) %>%
  na.omit() %>%
  filter(pop1!="Thai", pop2!="Thai", no_snps>=100)

pixy_fst$avg_wc_fst[pixy_fst$avg_wc_fst < 0] <- 0
pixy_fst$chromosome <- as.factor(pixy_fst$chromosome)
pixy_fst$midpos <- (pixy_fst$window_pos_1 + pixy_fst$window_pos_2)/2

# Load DXY data
pixy_dxy <- read_delim("./Data/win_slid/pixy_dxy.txt", delim="\t", trim_ws=TRUE) %>%
  na.omit() %>%
  filter(pop1!="Thai", pop2!="Thai", no_sites>=1000)

pixy_dxy$chromosome <- as.factor(pixy_dxy$chromosome)
pixy_dxy$midpos <- (pixy_dxy$window_pos_1 + pixy_dxy$window_pos_2)/2

# Deal with edge effect
trim_chromosome <- function(df) {
  df %>%
    group_by(chromosome) %>%
    group_modify(~ {
      d <- .x

      # Calculate mean and sd per chromosome
      sd_val <- sd(d$avg_dxy, na.rm = TRUE)
      mean_val <- mean(d$avg_dxy, na.rm = TRUE)
      threshold <- mean_val + sd_val 
      
      # Trimming in starting windows
      start_idx <- 1
      while (start_idx <= nrow(d) && d$avg_dxy[start_idx] > threshold) {
        start_idx <- start_idx + 1
      }
      
      #Trimming in ending windows
      end_idx <- nrow(d)
      while (end_idx >= start_idx && d$avg_dxy[end_idx] > threshold) {
        end_idx <- end_idx - 1
      }
      
      # Return trimmed dataset
      if (start_idx > end_idx) {
        # No valid windows left → return empty tibble
        return(d[0,])
      } else {
        return(d[start_idx:end_idx, ])
      }
    }) %>%
    ungroup()
}

pixy_dxy <- trim_chromosome(pixy_dxy)

# Scaling for secondary axis
b <- diff(range(pixy_fst$avg_wc_fst)) / diff(range(pixy_dxy$avg_dxy))
a <- min(pixy_fst$avg_wc_fst) - b * min(pixy_dxy$avg_dxy)

pixy_dxy$scaled_dxy <- a + b * pixy_dxy$avg_dxy

# Combine FST and DXY
combined <- bind_rows(
  pixy_fst %>% select(chromosome, midpos, value = avg_wc_fst) %>% mutate(type="FST"),
  pixy_dxy %>% select(chromosome, midpos, value = scaled_dxy) %>% mutate(type="DXY")
)

# For plot purpose, I transform the position so it starts at 0 and get the chromosome lengths to have different size of plots
combined <- combined %>%
  group_by(chromosome) %>%
  mutate(midpos0 = midpos - min(midpos)) %>%
  ungroup()

chr_lengths <- combined %>%
  group_by(chromosome) %>%
  summarise(chr_len = max(midpos0, na.rm=TRUE))

max_len <- max(chr_lengths$chr_len)

# Plot (one chromosome per row)
p <- ggplot(combined, aes(x = midpos0, y = value, color = type)) +
  geom_line(alpha = 0.5) +
  facet_grid(
    rows = vars(chromosome),
    scales = "fixed",
    switch = "y"
  ) +
  scale_x_continuous(limits = c(0, max_len)) +
  scale_y_continuous(
    "FST",
    sec.axis = sec_axis(~ (. - a)/b, name = "DXY")
  ) +
  labs(x = "Position along chromosome (normalized)") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.text.y.right = element_blank(),
    strip.background = element_blank()
  )
print(p)

# Search islands of differentiation (within the upper 2.5% of the distributions for both FST and DXY)
FST_only<-subset(combined, type=="FST")
high_FST<-subset(FST_only,  value>quantile(FST_only$value, 0.975))
high_FST<-high_FST[,-4]
colnames(high_FST)<-c("chromosome", "midpos", "FST", "midpos0")

DXY_only<-subset(combined, type=="DXY")
high_DXY<-subset(DXY_only,  value>quantile(DXY_only$value, 0.975))
high_DXY<-high_DXY[,-4]
colnames(high_DXY)<-c("chromosome", "midpos", "DXY", "midpos0")

all_high<-inner_join(high_FST, high_DXY)
all_high$start_window<-all_high$midpos-2500
all_high$end_window<-all_high$midpos+2500

# Only one region in chromosome 5 corresponded, so let's plot it
chr_5<-subset(combined, chromosome=="NC_041779.1")

p <- ggplot(chr_5, aes(x = midpos, y = value, color = type)) + 
  geom_rect(aes(xmin=1177501, xmax=1194001, ymin=0, ymax=0.1), fill="lightgrey", alpha=0.1, inherit.aes = FALSE)+
  geom_line(alpha = 0.5) +
  scale_y_continuous(
    "FST",
    sec.axis = sec_axis(~ (. - a)/b, name = "DXY")
  ) +
  labs(x = "Position along the chromosome") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.text.y.right = element_blank(),
    strip.background = element_blank()
  )
p
```

Does the window with significant differenciation signals fall into gene coding sequences (CDS)?
```{bash}
bedtools intersect -wb -a GCF_900090045.1_PmUG01_genomic.gff -b differenciation_region.tsv > gene_diff_ID.txt
```

## PBS

Version: pixy v2.0.0.beta14, R v4.5.2, bedtools v2.31.1.

Calculate *F<sub>ST</sub>* with pixy with sliding windows of 5kb and 500pb steps.
```{bash}
pixy --stats fst \
--vcf clusters_samples.filtered.vcf.gz \
--populations clusters_popfile.txt \
--bed_file slid.500stp.5kbwin.bed \
--n_cores 16
# the bed file corresponds to coordinates for sliding windows of 5kb and 500pb steps
```

Get *F<sub>ST</sub>* values and calculate PBS.
```{r}
library(tidyverse)
library(readr)

# Load FST values
pixy_fst <- read_delim("./Data/win_slid/pixy_fst.txt", delim="\t", trim_ws=TRUE) %>%
  na.omit() %>%
  filter(no_snps>=100)

pixy_fst$avg_wc_fst[pixy_fst$avg_wc_fst < 0] <- 0
pixy_fst$chromosome <- as.factor(pixy_fst$chromosome)

# Calculate PBS (T = -log(1 - fst))
pixy_fst <- pixy_fst %>%
  mutate(T = -log(1 - avg_wc_fst))

# Reshape the data so each window has 3 T values:
fst_wide <- pixy_fst %>%
  mutate(comparison = paste(pop1, pop2, sep = "_")) %>%
  select(chromosome, window_pos_1, window_pos_2, comparison, T) %>%
  pivot_wider(names_from = comparison, values_from = T)
fst_wide$midpos <- (fst_wide$window_pos_1 + fst_wide$window_pos_2)/2

# Now compute PBS for cluster Y and PBS for cluster R
fst_wide <- fst_wide %>%
  mutate(
    PBScY = (ClusterY_ClusterR + ClusterY_Thai - ClusterR_Thai) / 2,
    PBScR = (ClusterY_ClusterR + ClusterR_Thai - ClusterY_Thai) / 2
  )
fst_wide$PBScY[fst_wide$PBScY < 0] <- 0
fst_wide$PBScR[fst_wide$PBScR < 0] <- 0

# Preparation of the data for the manahattan plots
fst_wide$CHROM<-fst_wide$chromosome
fst_wide$chromosome<-as.numeric(as.factor(fst_wide$chromosome))
  data_cum <-  fst_wide %>% 
    group_by(chromosome) %>% 
    summarise(max_bp = max(midpos)) %>% 
    mutate(bp_add = lag(cumsum(max_bp), default = 0)) %>% 
    select(chromosome, bp_add)
   fst_wide <-  fst_wide %>% 
    inner_join(data_cum, by = "chromosome") %>% 
    mutate(bp_cum = midpos + bp_add)
  
  axis_set <-  fst_wide %>% 
    group_by(chromosome) %>% 
    summarize(center = max(bp_cum))

# Manhattan plot of Cluster Y
highlight<-subset(fst_wide,  PBScY>quantile(fst_wide$PBScY, 0.99))
write_tsv(as.data.frame(cbind(as.character(highlight$CHROM), highlight$window_pos_1, highlight$window_pos_2)), file = "./Data/PBScY_genes_signiff.tsv", col_names = F)

  p1<-fst_wide %>%
  sample_frac(0.5, replace = FALSE)  %>%
  ggplot(aes(x = bp_cum, y =PBScY,color = as_factor(chromosome))) +
    geom_point(alpha = 0.75) +
    scale_x_continuous(label = axis_set$chromosome, breaks = axis_set$center) +
    scale_color_manual(values = rep(c("#87CEEB", "#0047AB"), unique(length(axis_set$chromosome))))+
    labs(x = "Chromosome", y = "PBS") + 
    theme_minimal() +
    theme( 
      legend.position = "none",
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      axis.text.x = element_text())
  p1
p2<-p1+geom_point(data=highlight, aes(x = bp_cum, y =PBScY), color = "#bc4653", alpha= 0.8)
p2

p1<-fst_wide %>%
  sample_frac(1, replace = FALSE)  %>%
  ggplot(aes(x = bp_cum, y =PBScY,color = as_factor(chromosome))) +
    geom_point(alpha = 0.75) +
    scale_x_continuous(label = axis_set$chromosome, breaks = axis_set$center) +
    scale_color_manual(values = rep(c("#87CEEB", "#0047AB"), unique(length(axis_set$chromosome))))+
    labs(x = "Chromosome", y = "PBS") + 
    theme_minimal() +
    theme( 
      legend.position = "none",
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      axis.text.x = element_text())
  p1
p2<-p1+geom_point(data=highlight, aes(x = bp_cum, y =PBScY), color = "#bc4653", alpha= 0.8)
p2

# Manhattan plot of Cluster R
highlight<-subset(fst_wide,  PBScR>quantile(fst_wide$PBScR, 0.99))
write_tsv(as.data.frame(cbind(as.character(highlight$CHROM), highlight$window_pos_1, highlight$window_pos_2)), file = "./Data/PBScR_genes_signiff.tsv", col_names = F)


  p1<-fst_wide %>%
  sample_frac(0.5, replace = FALSE)  %>%
  ggplot(aes(x = bp_cum, y =PBScR,color = as_factor(chromosome))) +
    geom_point(alpha = 0.75) +
    scale_x_continuous(label = axis_set$chromosome, breaks = axis_set$center) +
    scale_color_manual(values = rep(c("#87CEEB", "#0047AB"), unique(length(axis_set$chromosome))))+
    labs(x = "Chromosome", y = "PBS") + 
    theme_minimal() +
    theme( 
      legend.position = "none",
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      axis.text.x = element_text())
  p1
p2<-p1+geom_point(data=highlight, aes(x = bp_cum, y =PBScR), color = "#bc4653", alpha= 0.8)
p2

p1<-fst_wide %>%
  sample_frac(1, replace = FALSE)  %>%
  ggplot(aes(x = bp_cum, y =PBScR,color = as_factor(chromosome))) +
    geom_point(alpha = 0.75) +
    scale_x_continuous(label = axis_set$chromosome, breaks = axis_set$center) +
    scale_color_manual(values = rep(c("#87CEEB", "#0047AB"), unique(length(axis_set$chromosome))))+
    labs(x = "Chromosome", y = "PBS") + 
    theme_minimal() +
    theme( 
      legend.position = "none",
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      axis.text.x = element_text())
  p1
p2<-p1+geom_point(data=highlight, aes(x = bp_cum, y =PBScR), color = "#bc4653", alpha= 0.8)
p2
```

Do windows with significant selection signals fall into gene coding sequences (CDS)?
```{bash}
bedtools intersect -wb -a GCF_900090045.1_PmUG01_genomic.gff \
-b PBScY_genes_signiff.tsv PBScR_genes_signiff.tsv \
-names PBScY PBScR > gene_ID.txt
```