---
title: "Worldwide population structure"
author: "Margaux Lefebvre"
date: "`r Sys.Date()`"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE) #prevent all chunks from running
library(tidyverse)
```

# Global PCA and ancestry plots

## ANGSD input and LD-pruning

Version: angsd v0.940, python v3.8.12, R v4.5.2, ngsLD v1.2.0.
```{bash}
listbam=all_filtered.noout.bamspaths #all samples without outgroup
REF=GCF_900090045.1_PmUG01_genomic.fna
CORE=core_malariae.txt

# Create Beagle input for the pruning
angsd -b $listbam -ref $REF -out PCA_coregenome \
        -rf $CORE -uniqueOnly 1 \
        -minMapQ 20 -minQ 20 -doCounts 1 \
        -GL 2 -doGlf 2 -nThreads 8 -doMajorMinor 1 -SNP_pval 1e-6 -doMaf 1 -minMaf 0.01

# Prepare a pos file with the mafs filre
zcat PCA_coregenome.mafs.gz | cut -f 1,2 |  sed 's/:/_/g'| gzip > PCA_coregenome.pos.gz

# Run ngsLD
ngsLD \
--geno PCA_coregenome.beagle.gz \
--posH PCA_coregenome.pos.gz \
--probs \
--n_ind 179 \
--n_sites 272444 \
--max_kb_dist 1 \
--n_threads 8 \
--out PCA_coregenome.ld

# Run prune_graph
prune_graph --header \
--in PCA_coregenome.ld --weight-field "r2" \
--weight-filter "dist <= 5000 && r2 >= 0.5" \
--out PCA_coregenome.snp.unlinked.id
```

Generate an LD-pruned SNP list:

```{r}
pruned_position <- as.integer(gsub(paste0("NC_0417[0-9][0-9].1:"), "", readLines(paste0("./Data/PCA_coregenome.snp.unlinked.id"))))
snp_list <- read.table(paste0("./Data/PCA_coregenome.mafs.gz"), stringsAsFactors = F, header = T)[,1:4]
pruned_snp_list <- snp_list[snp_list$position %in% pruned_position, ]
write.table(pruned_snp_list, paste0("./Data/PCA_coregenome.snp.LDpruned.list"), col.names = F, row.names = F, quote = F, sep = "\t")
```

Create the input for PCAngsd
```{bash}
listbam=all_filtered.noout.bamspaths #all samples without outgroup
REF=GCF_900090045.1_PmUG01_genomic.fna
CORE=core_malariae.txt

# Create input
angsd sites index PCA_coregenome.snp.LDpruned.list # mandatory

# Create Beagle input for the pruning
angsd -b $listbam -ref $REF -out PCA_coregenome_total \
        -rf $CORE -uniqueOnly 1 \
        -minMapQ 20 -minQ 20 -doCounts 1 \
        -GL 2 -doGlf 2 -nThreads 8 -doMajorMinor 1 -doMAF 1 -SNP_pval 1e-6 \
        -doPost 1 -doIBS 1 -doCov 1 -makeMatrix 1 -sites PCA_coregenome.snp.LDpruned.list
```

## PCA

Version: PCAngsd v1.36.1.
```{bash}
pcangsd -b PCA_coregenome_total.beagle.gz -o PCA_coregenome.05.mat --iter 500 -t 8 --maf 0.01
```

## Ancestry plots

Version: PCAngsd v1.36.1.

According to Meisner and Albrechtsen, the best K is determined by 1+ D (the optimal number of principal components). So D would be equal to 5 (determined by the elbow (broken-stick) method), so K=6.
```{bash}
# Run PCAngsd for ancestry plots
for k in {2..10}
do
 pcangsd -b PCA_coregenome_total.beagle.gz --admix --admix-K $k -o ancestry --iter 500 -t 8 --maf 0.01
done

# Run pong for vizualisation
pong -m filemap_tabs.txt -i ind2pop.txt -n pop_order.txt
```

# Maximum likelihood tree

Version: python v3.12, iqtree v2.3.4.
```{bash}
## Create the input
python vcf2phylip.py -i malariae_brasi.filt.core.ploidy2.vcf.gz -o GA02 --output-prefix tree_allsamples # change vcf to phylip file
## ML tree
iqtree -s tree_all.varsites.phy -m MFP+ASC -T 24 --prefix tree_all -o GA02 -B 1000 -alrt 1000 -st DNA #MFP+ASC = model finder for dataset with only variable sites
```